Algorithm explanation

### 2013-05-02 article/algo.tex
Update for constraint * Renaming \theta is used for index mapping

The Horn clause solving algorithm receives an input problem in Horn
graph. A Horn graph is a labelled directed graph $G=(V,E,\varphi)$.
\begin{itemize}
\item $V$ is a union of two sets of vertices; Horn term vertices $V_T$
  and arrow vertices $V_\Rightarrow$. Each vertex $u \in
  V_\Rightarrow$ have eactly one outgoing edge to some $v \in V_T$,
  denoted as $succ(u)$.
\item $E$ is a set of labelled edges, which are expressed in the form
  $(u,v,\theta)$ where $\theta$ is a finite map from variables to
  variables.
\item $\varphi: V_T \rightarrow L$ is a labelling function for Horn
  term vertices, where L is a set of elements of a predicate variable
  in the form $P(x_1, ..., x_{arity(P)})$ or a linear inequality $a_1
  x_1 + \cdots + a_n x_n \leq b$.
\end{itemize}

There are two kinds of edges in a Horn graph;
\begin{itemize}
\item for $u \in V_T, v \in V_\Rightarrow$, an edge takes a form
  $(u,v,\theta)$, or,
\item for $u \in V_\Rightarrow, v \in V_T$, an edge takes a form
  $(u,v,\emptyset)$, which the mapping is empty.
\end{itemize}

@@

We implicitly assume each $\mathcal{HC}_v$ is universally quantified
by its all free variables $\textsc{FV}(\mathcal{HC}_v)$.

@@

The solution of $G$ is a map $\rho$ from predicate variables to linear
predicates of the form $\lambda x_1, \cdots ,x_n. \psi $ (where $\psi$
is a formula constructed from
\begin{align*}
\psi ::= & a_1 x_1 + \cdots + a_n x_n \leq b \mid \\
& \psi \wedge \psi \mid \psi \vee \psi \mid true \mid false
\end{align*}
), where $\rho[G]$ is a tautology. A solution is \textit{simple} if
$\rho[P]$ is of the form $a_1 x_1 + \cdots + a_n x_n \leq b$ for every
$P$ in $dom(\rho)$.

@@

Given an acyclic Horn graph (or called a Horn DAG), we define the
relation $u <^+ v$ over $V_T$ as follows.
\[ u <^+ v \Longleftrightarrow \exists a, \theta; \left\lbrace (u,a,\theta), (a,v,\emptyset) \right\rbrace \subseteq E \]
By the assumption that $G$ is acyclic, $<^+$ is a well-founded
relation. We assume without loss of generality that $V_T$ contains the
greatest element $v_\bot$ with respect to $<^+$; otherwise we can
always implicitly assume $\varphi(v_\bot) = 0 \leq 0$ such that $\forall v \in V_\bot; \exists v_a; \left\lbrace (v, v_a, \emptyset), (v_a, v_\bot, \emptyset) \right\rbrace \subseteq E $.

@@

$\Delta(v)$ is defined by well-founded induction on $v$ with respect
to the relation $<^+$. Then, $C_v$ in
$\Delta(v) = \left( \mathbf{a}_v \mathbf{x}_v \leq b_v, C_v \right)$
can be shown as follows.

\begin{align*}
C_v = & \bigcup_{u <^+ v} C_u
\\
& \cup \begin{cases}
\emptyset
& \mbox{if } \nexists u; u <^+ v \\
\left\lbrace
 \mathbf{a}_v \mathbf{x}_v = \sum_{u <^+ v} \mathbf{a}_u \theta_{u,v} (\mathbf{x}_u),
 b_v \geq \sum_{u <^+ v} b_u
\right\rbrace
& \mbox{otherwise}
\end{cases}
\\
& \cup \begin{cases}
\emptyset
& \mbox{if } \varphi(v) = P(\mathbf{x}) \\
\left\lbrace
 \lambda_v \geq 0, \mathbf{a}_v = \lambda_v \mathbf{a}_0,
 b_v = \lambda_v b_0
\right\rbrace
& \mbox{if } \varphi(v) = \mathbf{a}_0 \mathbf{x} \leq b_0
\end{cases}
\end{align*}

For simplicity, we assume variables $\mathbf{a}_u$, $\mathbf{x}_u$,
$b_u$ are obtained from $\Delta(u)$ in each $\sum_{u <^+ v}$. Also, a
renaming function $\theta_{u,v}$ is shorthand for $\theta$ such that
$\exists a; \left\lbrace (u,a,\theta), (a,v,\emptyset) \right\rbrace
\subseteq E$, and here we pass a variable vector to it.

@@

\begin{quote}
If $\models \sigma(C_{v_\bot})$, then $\rho$ defined by
\begin{align*}
 \rho = \left\lbrace
  \left( P, \sigma(\mathbf{a}_v \mathbf{x} \leq b_v) \right) \middle|
  \forall v \in V; \varphi(v) = P(\mathbf{x})
 \right\rbrace
\end{align*}
is a solution of $G$.
\end{quote}

If $C_{v_\bot}$ is unsatisfiable, there are two possibilities for them.

\begin{itemize}
\item No solution exists for $G$.
\item No simple solution exists for $G$.
\end{itemize}

If $G$ has a vertex which has multiple successors, the second case is
possible. We now try to give a solution by relaxing the constraint
$C_{v_\bot}$.

@@

\begin{align*}
C_v = & \bigcup_{\substack{u <^+ v \\ u \not\in \hat V}} C_u \cup
\bigcup_{\substack{u <^+ v \\ u \in \hat V}} \textsc{Copy}_v (C_u)
\\
& \cup \begin{cases}
\emptyset
& \mbox{if } \nexists u; u <^+ v \\
\left\lbrace
 \mathbf{a}_{v,\emptyset} =
  \sum_{\substack{u <^+ v \\ u \not\in \hat V}} \mathbf{a}_{u,U} +
  \sum_{\substack{u <^+ v \\ u \in \hat V}} \mathbf{a}_{u,U \cup \left\lbrace v \right\rbrace},
 b_{v,\emptyset} \geq
  \sum_{\substack{u <^+ v \\ u \not\in \hat V}} b_{u,U} +
  \sum_{\substack{u <^+ v \\ u \in \hat V}} b_{u,U \cup \left\lbrace v \right\rbrace}
\right\rbrace
& \mbox{otherwise}
\end{cases}
\\
& \cup \begin{cases}
\emptyset
& \mbox{if } \varphi(v) = P(\mathbf{x}) \\
\left\lbrace
 \lambda_v \geq 0, \mathbf{a}_{v,\emptyset} = \lambda_v \mathbf{a}_0,
 b_{v,\emptyset} = \lambda_v b_0
\right\rbrace
& \mbox{if } \varphi(v) = \mathbf{a}_0 \mathbf{x} \leq b_0
\end{cases}
\end{align*}

@@

All variables are then distinguished by the label of copy origin
nodes. This intuitively means that constraints generated for nodes in
$\hat V$ are duplicated at their successors. Note that the variables
without the label should be interpreted as that it has $\emptyset$ for
origin nodes.

@@

\begin{align*}
 \rho = \left\lbrace
  \left( P, \bigwedge_S \sigma(\mathbf{a}_{v,S} \mathbf{x} \leq b_{v,S}) \right) \middle|
  \forall v \in V; \varphi(v) = P(\mathbf{x})
 \right\rbrace
\end{align*}

>>>>

The Horn clause solving algorithm receives an input problem in Horn
graph. A Horn graph is a labelled directed graph $G=(V,E,\varphi)$.
\begin{itemize}
\item $V$ is a union of two sets of vertices; Horn term vertices $V_T$
  and arrow vertices $V_\Rightarrow$. Each vertex $u \in
  V_\Rightarrow$ has eactly one outgoing edge to some $v \in V_T$,
  denoted as $succ(u)$.
\item $E$ is a set of labelled edges, which are expressed in the form
  $(u,v,\theta)$ where $\theta$ is a finite map from variables to
  variables.
\item $\varphi: V_T \rightarrow L$ is a labelling function for Horn
  term vertices, where L is a set of elements of a predicate variable
  in the form $P(x_1, ..., x_{arity(P)})$ or a linear inequality $a_1
  x_1 + \cdots + a_n x_n \leq b$.
\end{itemize}

There are two kinds of edges in a Horn graph;
\begin{itemize}
\item for $u \in V_T, v \in V_\Rightarrow$, an edge takes a form
  $(u,v,\theta)$, or,
\item for $u \in V_\Rightarrow, v \in V_T$, an edge takes a form
  $(u,v,\emptyset)$, where the mapping is empty.
\end{itemize}

@@

We implicitly assume each $\mathcal{HC}_v$ is universally quantified
by its all free variables $\textsc{FV}(\mathcal{HC}_v)$. Without loss
of generality, we limit occurrence of linear expression on right-hand
side of Horn clauses to either $0 \leq -1$, meaning $\bot$, or $0 \leq
0$, meaning $\top$.

@@

The solution of $G$ is a map $\rho$ from predicate variables to linear
predicates of the form $\lambda x_1, \cdots ,x_n. \psi $ (where $\psi$
is a formula constructed from
\begin{align*}
\psi ::= & a_1 x_1 + \cdots + a_n x_n \leq b \mid \\
& \psi \wedge \psi \mid \psi \vee \psi \mid \top \mid \bot
\end{align*}
), where $\rho[G]$ is a tautology. A solution is \textit{simple} if
$\rho[P]$ is of the form $a_1 x_1 + \cdots + a_n x_n \leq b$ for every
$P$ in $dom(\rho)$.

@@

For simplicity of discussion, we define $k_v$ the arity of the
right-hand side of $\mathcal{HC}_v$. Assuming that variables in every
Horn clauses are appropriately renamed, the right-hand side of
$\mathcal{HC}_v$ takes the form $P(x_1,...,x_{k_v})$ where $k_v =
arity(P)$, or either $\top$ or $\bot$ with $k_v = 0$. Also we let $K$
denote the maximum of $k_v$ over all $v$.

Given an acyclic Horn graph (or called a Horn DAG), we define the
relation $u \xrightarrow{\theta} v$ over $V_T$ as follows.
\[ u \xrightarrow{\theta} v \Longleftrightarrow
\exists a \in V_\Rightarrow; \left\lbrace (u,a,\theta), (a,v,\emptyset) \right\rbrace \subseteq E \]

We may omit $\theta$ and write $u \rightarrow v$ for convenience for
later discussion. By the assumption that $G$ is acyclic,
$\rightarrow^+$ is a well-founded relation. We assume without loss of
generality that $V_T$ contains the greatest element $v_\bot$ with
respect to $\rightarrow^+$; otherwise we can always implicitly assume
$\varphi(v_\bot) = \top$ such that $\forall v \in V_T; v
\xrightarrow{\emptyset} v_\bot$.

@@

$\Delta(v)$ is defined by well-founded induction on $v$ with respect
to the relation $\rightarrow^+$. Then, $C_v$ in
$\Delta(v) = \left( \mathbf{a}_v \mathbf{x}_v \leq b_v, C_v \right)$
can be shown as follows.

\begin{align*}
C_v = & \bigcup_{u \rightarrow v} C_u
\\
& \cup \begin{cases}
\emptyset
& \mbox{if } \nexists u; u \rightarrow v \\
\left\lbrace
 \mathbf{a}_{v,i} = \sum_{u \xrightarrow{\theta} v} \mathbf{a}_{u, \theta^{-1} (i)},
 b_v \geq \sum_{u \xrightarrow{\theta} v} b_u
\right\rbrace \cup \\
\hspace{2cm} \bigcup_{k_v < i \leq K} \left\lbrace \mathbf{a}_{v,i} = 0 \right\rbrace
& \mbox{otherwise}
\end{cases}
\\
& \cup \begin{cases}
\emptyset
& \mbox{if } \varphi(v) = P(\mathbf{x}_v) \\
\left\lbrace
 \lambda_v \geq 0, \mathbf{a}_v = \lambda_v \mathbf{a}_0,
 b_v = \lambda_v b_0
\right\rbrace
& \mbox{if } \varphi(v) = \mathbf{a}_0 \mathbf{x}_v \leq b_0
\end{cases}
\end{align*}

For simplicity, we assume variables $\mathbf{a}_u$, $b_u$ are obtained
from $\Delta(u)$.

@@

\begin{quote}
If $\models \sigma(C_{v_\bot})$, then $\rho$ defined by
\begin{align*}
 \rho = \left\lbrace
  \left( P, \sigma(\mathbf{a}_v \mathbf{x}_v \leq b_v) \right) \middle|
  \forall v \in V_T; \varphi(v) = P(\mathbf{x}_v)
 \right\rbrace
\end{align*}
is a solution of $G$.
\end{quote}

If $C_{v_\bot}$ is unsatisfiable, no simple solution exists for
$G$. We now try to give a solution by relaxing the constraint
$C_{v_\bot}$.

@@

\begin{align*}
C_v = & \bigcup_{\substack{u \rightarrow v \\ u \not\in \hat V}} C_u \cup
\bigcup_{\substack{u \rightarrow v \\ u \in \hat V}} \textsc{Copy}_v (C_u)
\\
& \cup \begin{cases}
\emptyset
& \mbox{if } \nexists u; u \rightarrow v \\
\left\lbrace
 \mathbf{a}_{v,\emptyset,i} =
  \sum_{\substack{u \xrightarrow{\theta} v \\ u \not\in \hat V}} \mathbf{a}_{u,U,\theta^{-1}(i)} +
  \sum_{\substack{u \xrightarrow{\theta} v \\ u \in \hat V}}
  \mathbf{a}_{u,U \cup \left\lbrace v \right\rbrace, \theta^{-1}(i)}
\right\rbrace \cup \\
\hspace{2cm} \left\lbrace
 b_{v,\emptyset} \geq
  \sum_{\substack{u \rightarrow v \\ u \not\in \hat V}} b_{u,U} +
  \sum_{\substack{u \rightarrow v \\ u \in \hat V}} b_{u,U \cup \left\lbrace v \right\rbrace}
\right\rbrace \cup \\
\hspace{2cm} \bigcup_{k_v < i \leq K} \left\lbrace \mathbf{a}_{v,\emptyset,i} = 0 \right\rbrace
& \mbox{otherwise}
\end{cases}
\\
& \cup \begin{cases}
\emptyset
& \mbox{if } \varphi(v) = P(\mathbf{x}_v) \\
\left\lbrace
 \lambda_v \geq 0, \mathbf{a}_{v,\emptyset} = \lambda_v \mathbf{a}_0,
 b_{v,\emptyset} = \lambda_v b_0
\right\rbrace
& \mbox{if } \varphi(v) = \mathbf{a}_0 \mathbf{x}_v \leq b_0
\end{cases}
\end{align*}

@@

All variables are then distinguished by the label of copy origin
nodes. This intuitively means that constraints generated for nodes in
$\hat V$ are duplicated at their successors. For consistency, the
variables without the label should be interpreted as that it has
$\emptyset$ for origin nodes.

@@

\begin{align*}
 \rho = \left\lbrace
  \left( P, \bigwedge_S \sigma(\mathbf{a}_{v,S} \mathbf{x}_v \leq b_{v,S}) \right) \middle|
  \forall v \in V_T; \varphi(v) = P(\mathbf{x}_v)
 \right\rbrace
\end{align*}
